---
title: "T-cell Acute Lymphoblastic Leukemia RNA Seq Analysis "
author: "Ling Min Hao"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false 
    code_folding: hide
    theme: journal
    highlight: haddock
bibliography: references.bib 
nocite: '@*'
---

```{r setup and library, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
# Setup
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      cache = FALSE, echo = TRUE)
```

```{r Libraries, results = FALSE,  warning = FALSE, message = FALSE}
# Load the libraries
library(DT)
library(recount)
library(DESeq2)
library(tidyverse)
library(tibble)
library(EnhancedVolcano)
library(knitr)
library(plyr)
library(EnsDb.Hsapiens.v86)
library(pheatmap)
library(msigdbr)
library(clusterProfiler)
library(ggpubr)
library(rlist)
```

# Introduction

**T-cell acute lymphoblastic leukemia (T-ALL)** is a specific type of leukemia. **BBC3**
is the gene helpful to inhibit the growth and proliferation of T-ALL, and **HES1** in the
**NOTCH1 signaling pathway** downregulates BBC3. On the other hand, perhexiline has some
good antileukemic effect on T-ALL.[@schnell2015therapeutic]. 

So, we proposed that perhexiline might be a drug that downregulates HES1. To address this
gap in knowledge, we performed differential gene expression (DGE) analysis on data from
CUTLL1, a novel human T-cell lymphoma cell lines. We performed the analysis on
**three control samples** and **three samples treated with perhexiline**, an antagonist 
drug with robust anti-leukemic activity against **T-ALL** in vitro and in vivo using the 
[dataset SRP055108](https://trace.ncbi.nlm.nih.gov/Traces/sra/?study=SRP055108).

```{r DESeq dataset}
### Prepare DEseq dataset
## A leukemia dataset is selected and loaded from recount 2
selected_study <- "SRP055108"
if (! file.exists("SRP055108/rse_gene.Rdata")){
  download_study(selected_study)
} 
load("SRP055108/rse_gene.Rdata")

## Tidy the dataset 
# Convert ENSG to Gene Symbol 
ens2sym <- AnnotationDbi::select(EnsDb.Hsapiens.v86, keys = keys(EnsDb.Hsapiens.v86),
                                 columns = c("SYMBOL"))
# Count Matrix
countData <- as.data.frame(assay(rse_gene)) %>% 
  rownames_to_column("GENEID") %>% 
  filter(!grepl(pattern = "_PAR_Y",GENEID)) %>%  
  mutate(GENEID = gsub(GENEID, pattern = "\\..+", replacement = "")) %>% # cleaning the ensembl id
  inner_join(ens2sym, by = "GENEID") %>%  # connect with gene symbol
  dplyr::select(-GENEID) %>% 
  group_by(SYMBOL) %>% 
  summarise(across(everything(), ~ sum(., is.na(.), 0))) %>% # aggregrate rows with same gene symbol
  ungroup() %>% 
  column_to_rownames("SYMBOL") # count matrix with row name (gene symbol) and column name (samples)

# Assign the label "Control" or "Treated" to the samples
colData(rse_gene)$condition <- gsub(colData(rse_gene)$title, pattern = "^C.*",
                                    replacement = "Control")
colData(rse_gene)$condition <- gsub(colData(rse_gene)$condition, pattern = "^T.*",
                                    replacement = "Treated")
colData <- as.data.frame(colData(rse_gene)) %>% 
  dplyr::select(condition)

## Convert to DESeq dataset 
dds <- DESeqDataSetFromMatrix(countData, colData, design = ~condition) 
```

```{r sample and condition}
# Table showing controlled and treated samples 
datatable(colData, colnames = c("Sample", "Condition"), 
          caption = "Control and Treated Samples from CUTLL cell lines")
```

We will first need to investigate the quality of dataset. The PCA plot (Figure \@ref(fig:QC))
below shows that most variation in the dataset is biological. So, this is a good dataset. 

```{r QC, fig.align = "center", fig.cap = "PCA Plot. 78% of the variation of the dataset are due to biological variation"}
## Quality Check of dataset
# Get rlog 
rld <- rlog(dds)
# plot PCA
plotPCA(rld)
```

# Results 
## Differential Gene Expression (DGE) Analysis

```{r DESeq2}
## DESeq2 Analysis 
# Get results 
dds <- DESeq(dds)
res <- results(dds, contrast = c("condition", "Treated", "Control"))
# LFC shrink (to compensate for low count genes with high dispersion values)
resNorm <- lfcShrink(dds = dds, res = res, type = "normal", coef = 2)
# Make into data frames
resdf <- as.data.frame(resNorm) %>% rownames_to_column("SYMBOL")
```

We first performed differential gene expression analysis. Then, we obtained the volcano
plot (Figure \@ref(fig:VolcanoPlot)) visualizing the amount of differentially expressed
genes (DEG) uncovered

```{r VolcanoPlot, fig.height = 7, fig.width = 9, fig.align = "center", fig.cap = "Volcano plot (P < 0.05) visualizing the amount of DEGs after DGE Analysis. The most overexpressed genes are towards the right, the most underexpressed genes are towards the left, and the most statistically significant genes are towards the top."}
EnhancedVolcano(resdf, 
                lab = resdf$SYMBOL,
                pCutoff = 0.05,
                x = "log2FoldChange", 
                y = "padj",
                title = "Differentially Expressed Genes (DEG)", 
                subtitle = "", 
                legendLabels=c('Not sig.','Log (base 2) FC','p-value', 'p-value & Log (base 2) FC'), 
                pointSize = 4.0,
                labSize = 6.0)
```

Here we list out all differentially expressed genes. 

```{r DEGlist, results="asis"}
# All over-expressed and under-expressed gene list
deg_list <- resdf %>% 
  dplyr::filter(padj < .05) %>% # Filter for significant results
  mutate(result = case_when(
    log2FoldChange > 0 ~ "Over-expressed", 
    TRUE ~ "Under-expressed"
  )) %>% 
  group_by(result) %>% # Group by result column 
  arrange(padj) %>% 
  {setNames(group_split(.), group_keys(.)[[1]])} # Split tibble into list by group with names

# Tabulate the gene list
GENE <- llply(names(deg_list), function(groupNow){
  datatable(deg_list[[groupNow]] %>% dplyr::select(c(SYMBOL, baseMean, log2FoldChange, padj)), 
            caption = paste0(groupNow, ' Gene List'), 
            )
})

GENE[[1]]
GENE[[2]]
```

We now study the top 20 overexpressed and underexpressed DEGs using heatmap (Figure \@ref(fig:Heatmap)) below. 

```{r Heatmap, fig.height = 7, fig.width = 9, fig.align = "center", fig.cap = "Heat map representation of the top 20 differentially under-expressed and over-expressed genes (P < 0.05) between control and perhexiline-treated CUTLL1 cells. The scale bar shows color-coded differential gene expression, with red indicating high gene expression levels and blue indicating lower gene expression levels."}
# Get top 20 over-expressed and under-expressed gene list 
top20_deg_list <- deg_list %>% 
  llply(slice_min, order_by = padj, n = 20)
# Subset count matrix from top 20 over-expressed and top 20 under-expressed genes
top20_degs_mat <- as.data.frame(assay(rld)) %>%
  dplyr::filter(rownames(assay(rld)) %in% 
                  (rbind(top20_deg_list$`Over-expressed`, top20_deg_list$`Under-expressed`) %>% 
                  dplyr::pull(SYMBOL)))

# Plot the heatmap
pheatmap(top20_degs_mat, scale = "row", 
         clustering_distance_rows = "correlation", annotation_col = colData,
         main="Top 20 Differentially Underexpressed and Overexpressed Genes")
```

## Pathway Analysis 
### Broad-Spectrum Pathway Analysis{.tabset}

Using the overexpressed and underexpressed genes, we perform a brief pathway analysis using
[Enrichr](https://maayanlab.cloud/Enrichr/). The holistic results are shown in the link below. 

```{r childRMD, results='asis'}
# Get the gene sets and wrangle
gene_sets <- msigdbr(species = "Homo sapiens", category = "C5")
gene_sets <- gene_sets %>%
  dplyr::select(gs_name, gene_symbol)

# Pull the gene symbol from DEGs
deg_symbol <- deg_list %>% 
  llply(pull, var = "SYMBOL") 

ENRICHR <- llply(names(deg_symbol), function(groupNow) {
  genesNow <- deg_symbol[[groupNow]]
  response <- httr::POST(  # Post request to enrichr based on https://maayanlab.cloud/Enrichr/help#api&q=1
    url = 'https://maayanlab.cloud/Enrichr/addList', 
    body = list(
      'list' = paste0(genesNow, collapse = "\n"),
      'description' = groupNow
      )
    )
  response <- jsonlite::fromJSON(httr::content(response, as = "text"))  # Collect response
  permalink <- paste0("https://maayanlab.cloud/Enrichr/enrich?dataset=",  # Create permalink
                      response$shortId[1])
  # See this for more guidance: https://bookdown.org/yihui/rmarkdown-cookbook/child-document.html
  knitr::knit_child(text = c(  # Text vector to be knitted into RMarkdown as a child
    '#### `r groupNow` Pathways',
    '',
    'Enrichr Link: <a href="`r permalink`" target="_blank">`r groupNow`</a>.',
    ''
  ), 
  envir = environment(),  # Current global environment will be passed into the RMarkdown child
  quiet = TRUE)
})

cat(unlist(ENRICHR), sep = '\n')
```

### Gene Set Enrichment Analysis

We perform Gene Set Enrichment Analysis (GSEA) on the ontology gene sets to obtain a ranked 
list of upregulated and downregulated biological pathways on perhexiline treatment in CUTLL1 cells. 

```{r GSEA, results = "asis"}
# Get the gene sets and wrangle
gene_sets <- msigdbr(species = "Homo sapiens", category = "C5")
gene_sets <- gene_sets %>%
  dplyr::select(gs_name, gene_symbol)

# Adding a score for GSEA
resdf2 <- resdf %>%
  arrange(padj) %>%
  mutate(gsea_metric = -log10(padj) * sign(log2FoldChange))

# Remove NAs and order by GSEA
resdf2 <- resdf2 %>%
  filter(! is.na(gsea_metric)) %>%
  arrange(desc(gsea_metric))

# Get the ranked GSEA vector
ranks <- resdf2 %>%
  select(SYMBOL, gsea_metric) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  deframe()

# Run GSEA
gseares <- GSEA(geneList = ranks, 
                TERM2GENE = gene_sets)
gsearesdf <- as.data.frame(gseares)

pathway_list <- gsearesdf %>% 
  mutate(result = case_when(
    NES > 0 ~ "Over-expressed", 
    TRUE ~ "Under-expressed"
  )) %>% 
  group_by(result) %>% # Group by result column 
  arrange(desc(abs(NES))) %>% # Arrange the rows according to NES score
  {setNames(group_split(.), group_keys(.)[[1]])} # Split tibble into list by group with names

# Tabulate the pathway list
PATHWAY <- llply(names(pathway_list), function(groupNow){
  datatable(pathway_list[[groupNow]] %>% dplyr::select(c(ID,NES, p.adjust)), 
            caption = paste0(groupNow, ' Biological Pathways'), 
            )
})

PATHWAY[[1]]
PATHWAY[[2]]
```

Specifically, we analyze the top 5 upregulated and downregulated biological pathways on 
perhexiline treatment in CUTLL1 cells. 

```{r GSEAPlot, results = "asis", fig.width = 32, fig.height = 20, fig.align = "center", fig.cap = c("GSEA analysis plots of top 5 upregulated biological pathways on perhexiline treatment in CUTLL1 cells", "GSEA analysis plots of top 5 downregulated biological pathways on perhexiline treatment in CUTLL1 cells")}
## Make GSEA plot for top and bottom results
# Get top 5 over-expressed and under-expressed biological pathways 
top5_pathway_list <- pathway_list %>% 
  llply(slice_max, order_by = abs(NES), n = 5) 
## -- Make gseaplot for each and return as list
TOPPATHWAY <- llply(names(top5_pathway_list), function(groupNow){
  list <- top5_pathway_list[[groupNow]] %>% pull(ID)
  top_pathway_plots <- lapply(list, function(pathway) {
    gseaplot(gseares, geneSetID = pathway, title = pathway)
  })
  ## -- Arrange with labels as a multi-panel plot
  final <- ggarrange(plotlist = top_pathway_plots, ncol = 3, nrow = 2, labels = "AUTO")
  annotate_figure(final, top = text_grob(paste0("Top 5 ", groupNow, " Biological Pathways"), face = "bold", size = 30))
})

TOPPATHWAY[[1]]
TOPPATHWAY[[2]]
```

# Discussion

Based on the analysis, perhexiline did not significantly downregulate HES1. In fact, 
perhexiline slightly upregulates HES1 (it is ranked 94 in the DEG gene list). So there are
other reasons why perhexiline treats T-ALL. After analyzing the biological pathway using
GSEA, most biological processes related to cell division and cell growth are downregulated. 
This is not a surprising result because the production of tumors is due to abniormal rapid 
cell growth. Since perhexiline stops T-ALL, we expect the downregulation of those processes. 

Yet, it is unknown whether perhexiline directly caused the downregulation. The finding shows
perhexiline upregulates most processes related to the production of steroids. So, it could 
be that steroid is the one that downregulated cell division and cell growth processes. 
As a result, steroids might be the potential treatment for T-ALL. Further investigation 
is needed to verify this new conjecture. 

# References 
